---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# grtsdb

<!-- badges: start -->
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://lifecycle.r-lib.org/articles/stages.html)
![GitHub](https://img.shields.io/github/license/inbo/grtsdb)
[![R build status](https://github.com/inbo/grtsdb/workflows/R-CMD-check/badge.svg)](https://github.com/inbo/grtsdb/actions)
[![Codecov test coverage](https://codecov.io/gh/inbo/grtsdb/branch/master/graph/badge.svg)](https://app.codecov.io/gh/inbo/grtsdb?branch=master)
![GitHub code size in bytes](https://img.shields.io/github/languages/code-size/inbo/grtsdb.svg)
![GitHub repo size](https://img.shields.io/github/repo-size/inbo/grtsdb.svg)
<!-- badges: end -->

The goal of `grtsdb` is to create a spatially balanced sample based on the 'Generalised Random Tesselation Stratified' strategy.
We store the base schema in an SQLite database to make the sampling reproducible.
Sampling the same database with the same parameters yields a stable sample.

## Installation

To install the latest stable version use

``` r
# activate the INBO r-universe
options(
  repos = c(
    INBO = 'https://inbo.r-universe.dev', CRAN = 'https://cloud.r-project.org'
  )
)
install.packages("grtsdb")
```

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("inbo/grtsdb")
```

## Example

This is a basic example.

Connect to a database.

```{r}
tmp_copy <- tempfile(pattern = "grts", fileext = ".sqlite")
if (system.file("grts.sqlite", package = "grtsdb") != "") {
  file.copy(system.file("grts.sqlite", package = "grtsdb"), tmp_copy)
}
library(grtsdb)
db <- connect_db(tmp_copy)
```

To extract a sample, you'll need to specify the bounding box in projected coordinates and the size of the grid cells.

```{r example}
bbox <- rbind(
  c(0, 32),
  c(0, 32)
)
extract_sample(grtsdb = db, samplesize = 10, bbox = bbox, cellsize = 1)
```

Repeating the sample yields the same results.

```{r}
extract_sample(grtsdb = db, samplesize = 10, bbox = bbox, cellsize = 1)
```

```{r echo = FALSE}
drop_legacy_sites(grtsdb = db, level = 5)
```

You can add legacy sites to the sampling scheme.

```{r}
legacy <- rbind(
  c(4, 4),
  c(17, 6)
)
add_legacy_sites(legacy, bbox = bbox, cellsize = 1, grtsdb = db)
extract_legacy_sample(grtsdb = db, samplesize = 10, bbox = bbox, cellsize = 1)
```

You can compact the database for storage.

```{r}
compact_db(db)
```

Disconnect the database when done.

```{r}
dbDisconnect(db)
```

```{r eval = system.file("grts.sqlite", package = "grtsdb") == "", echo = FALSE}
dir.create("inst", showWarnings = FALSE)
file.copy(tmp_copy, file.path("inst", "grts.sqlite"))
```
