---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# grtsdb

<!-- badges: start -->
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
![GitHub](https://img.shields.io/github/license/inbo/grtsdb)
[![R build status](https://github.com/inbo/grtsdb/workflows/R-CMD-check/badge.svg)](https://github.com/inbo/grtsdb/actions)
[![Codecov test coverage](https://codecov.io/gh/inbo/grtsdb/branch/master/graph/badge.svg)](https://codecov.io/gh/inbo/grtsdb?branch=master)
![GitHub code size in bytes](https://img.shields.io/github/languages/code-size/inbo/grtsdb.svg)
![GitHub repo size](https://img.shields.io/github/repo-size/inbo/grtsdb.svg)
<!-- badges: end -->

The goal of `grtsdb` is to create as statially balanced sample based on the 'Generalised Random Tesselation Stratified' strategy.
We store the base schema in an SQLite database.
Store this database to make the sampling reproducible.
Sampling the same database with the same parameters yields a stable sample.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("inbo/grtsdb")
```

## Example

This is a basic example.
For demonstration purposes, we store the database in memory.
In practice you should store the database on a hard drive an archive it.

```{r example}
library(grtsdb)
# connect database
db <- connect_db(":memory:")
# bounding box of the target area
bbox <- rbind(
  c(0, 100),
  c(0, 100)
)
extract_sample(grtsdb = db, samplesize = 20, bbox = bbox, cellsize = 1)
```
Repeating the sample yields the same results

```{r}
extract_sample(grtsdb = db, samplesize = 20, bbox = bbox, cellsize = 1)
```

You can add legacy sites to the sampling scheme.

```{r}
legacy <- rbind(
  c(25, 25),
  c(51, 12)
)
add_legacy_sites(legacy, bbox = bbox, cellsize = 1, grtsdb = db)
extract_legacy_sample(grtsdb = db, samplesize = 20, bbox = bbox, cellsize = 1)
```

Disconnect database when done.

```{r}
dbDisconnect(db)
```

